Synthesizing Quotient Lenses

### 2018-02-02 papers/QuotientLenses/main.tex
edit quotient lens paper

{\em Lenses}~\cite{Lenses} specify bidirectional 
transformations between different data formats.  Writing a single lens
rather than two separate uni-directional transformations saves both
keystrokes and bugs, by reducing redundancy and by automatically
guaranteeing strong ``lens laws'' ensuring that the forward and backward
transformations interact well.

{\em Quotient lenses}~\cite{quotientlenses} are a generalization of lenses
where each of the data formats is quotiented by some equivalence relation
and the lens laws are taken modulo equivalence.  This allows lenses to be
both simpler and more flexible, since whole equivalence classes of data
values can be handled all at once and inessential details like whitespace
and ordering of fields can be ignored.

However, while plain lenses are widely used in both standalone systems and
popular libraries, Boomerang~\cite{boomerang} is the only implementation
currently offering quotient lenses.  One reason for this is that there is
some annoying redundancy in the quotient lens notation: The programmer often
has to separately specify (1) the data formats on each side of the lens, (2) the
equivalence relations over them, and (3) a lens between the canonical
representatives of the equivalence classes, even though, typically, these three
share a great deal of structure.  Moreover, writing the lens between the
equivalence classes can be challenging due to the rather fiddly typechecking,
mainly due to ``non-ambiguity'' conditions in the lens well-formedness rules.

In this paper, we make programming quotient lenses easier in two ways. First, we
introduce a novel language of {\em Quotient Regular Expressions}, or QREs, which
enable a programmer to simultaneously specify a regular expression as well as an
equivalence relation on that regular expression. QREs make programming quotient
lenses easire by saving the programmer the effort of defining the data and the
equivalence relation on the data separately.

Secondly, we use QREs to define a class of quotient lenses called {\em QRE
lenses} which map betweeen QREs, and show how to synthesize QRE lenses given
an specification described by pair of QREs and a set of input-output examples.
This way, the programmer need not manually derive the lens herself. The net
effect of these two contributions, that is QREs and QRE lens synthesis, is that
in order to derive a quotient lens, the programmer will only need to specify the
data formats on each side of the desired quotient lens and possibly a set of
example input-output pairs that guide synthesis of the lens, a marked
improvement over the current situation.

In summary, our main contributions are:
\begin{enumerate}
\item We introduce a novel language of {\em Quotient Regular Expressions}
(QREs) that a provide compact, convenient notation for a natural and useful
class of equivalence relations on regular languages (Section~\ref{QRE}).
\item We define a language of {\em QRE lenses}, a subset of bijective quotient
lenses whose types are given by QREs. Our main technical contribution in this
paper is a proof that each QRE lens $q$ defines the same transformation as a
QRE lens $q'$ that is in normal form. (Section~\ref{QRE-lenses}).
\item Using this normal form, we reduce the problem of {\em synthesizing}
QRE lenses from their QRE types and a set of examples to the previously studied
problem of synthesizing bijective lenses between regular languages 
(Section~\ref{synth}).
\item We extend the Boomerang {\em implementation} with QREs, QRE lens operators,
and QRE lens synthesis and demonstrate its applicability by using it to
synthesize QRE lenses between several real-world data formats from the
{\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

\section{Quotient Regular Expressions and Lenses by Example}
\label{sec:example}
In this section, we explain the difficulties of programming with quotient lenses
and show, by way of example, how our novel language of {\em Quotient Regular
Expressions}, or QREs and QRE lenses address these difficulties.

@@

We wish for this transformation to have the same behaviour on records that
differ only in ``inessential'' details. First, the transformation
should allow whitespace characters to occur arbitrarily between the words in the
label, author and title fields as well, as in between them. Second, an
author's name can have the surname followed by a comma and then the other
names separated by arbitrary whitespace, or just all the names separated by
arbitrary whitespace (for example, the names ``Conway, J. H.'' and ``J. H.
Conway'' are equivalent). Lastly, the transformation should also ignore the
order in which the author and title fields occur in each citation record.

@@

While it is possible to express the equivalence on the \bibtex{} data above in
Boomerang, actually doing so is a challenge. This is because the entire
equivalence is composed of three separate equivalences---whitespace quotienting
inside fields, whitespace quotenting in between fields, and permutation
quotenting of fields---nest with each other in a way that makes them difficult
to express with the standard Boomerang combinators.

In general, there is also another issue in describing equivalences in
Boomerang, which is that that the programmer often has to first specify the data
itself in the form of a regular expression before she can define the
equivalence relation on that data. This leads to verbose code.

To address these issues, we propose {\em Quotient Regular Expressions} or QREs
which enable the programmer to define a regular expression $R$ as well as an
equivalence relation on $R$ in the same expression. For example, with QREs we
can specify \bibtex{} records as shown in Figure~\ref{fig:example-qre}.

\begin{figure}[t]
\begin{lstlisting}
let wsp = [ \n\t\r]

(* map arbitrary occurrences of whitespace to a single space *)
let wsp_to_space   = project wsp$^+$ $\to$ " "

(* define name representations with a space and with a comma *)
let space_name     = name | (name . wsp_to_space . (name . wsp_to_space)$^*$ . name)
let comma_name   = name | (name . "," . wsp_to_space . (name . wsp_to_space)$^*$ . name)

@@

(* squash QRE maps comma_name to space_name *)
let valid_name = squash comma_name $\to$ space_name using (get space_to_comma)

(* define rest of bibtex fields *)
let wsp_to_and = project (wsp$^+$ . "and" . wsp$^+$) $\to$ " and " 
let bib_names  = valid_name | (valid_name . wsp_to_and . (valid_name . wsp_to_and)$^*$ . valid_name) 
let bib_author = "author = \"" . bib_names . "\""
let title     = word | (word . wsp_to_space . (word . wsp_to_space)$^*$ . word) 
let bib_title = "title = {" . title . "}"

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_perm  = perm (label, bib_author, bib_title) with (project ("," .wsp$^+$) $\to$ ",\n")
let bibtex    = "@book{" . bib_perm . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. }
  \label{fig:example-qre}
\end{figure}

In Figure~\ref{fig:example-qre}, the QRE \cd{project wsp}+ $\to$ \cd{" "}
matches any non-zero number of whitespace characters, but considers the matched
string of whitespace characters as equivalent with a single space as the
canonical representative. We call this QRE a \textit{projection} because it
projects an entire regular language down to a single representative string in
the language.

@@

The third example of a QRE in the Figure is \cd{bib_perm}, defined by:
\begin{center}
\begin{lstlisting}
perm (label, bib_author, bib_title) with (project ("," . wsp$^+$ ) $\to$ ",\n")
\end{lstlisting}
\end{center}

\noindent This is a \textit{permutation} QRE that specifies that the label,
title, and author fields can occur in any order, with the fields separated by a
comma followed by arbitrary amounts of whitespace. The canonical representation
is that in which the label, author and title fields occur in that order,
separated by a comma followed a newline. The \textit{permutation} QREs is
designed for equivalences such as this, where different permutations of
equivalences are interspersed with a separator which itself has an
equivalence defined on it.

@@

Consequently, we offer the programmer some relief by identifying a class of
quotient lenses that we call {\em QRE lenses}, and show how to {\em synthesize}
QRE lenses from a pair of QREs and a set of input-output example pairs. Our key
insight is to reduce the problem of synthesizing QRE lenses to the problem of
synthesizing {\em bijective lenses}, which are the bijections created from the
Boomerang lens combinators. This is a problem which we solved in our prior work
~\cite{popl18}.

@@

This section introduces Quotient Regular Expressions or QREs. QREs are regular
expressions augmented with syntax that lets them simultaneously express an
equivalence relation on the language described by the regular expression.

\subsection{Syntax of Quotient Regular Expressions}
The language of Quotient Regular Expressions (QREs) is given by the following
grammar: 
\begin{align*}
c := \; &R \sep \project{R}{s} \sep \squash{c}{c'}{f} \sep
\perm{c_1, \ldots, c_n}{c} \\
  & | \; \normalize{R}{R'}{f} \sep c \semicolon c' \sep c \cdot c' \sep (c \sep
  c') \sep c^*,
\end{align*}
where $c$ ranges over QREs, $R$ ranges over regular expressions, $f$ ranges over
functions between regular languages, and $s$ ranges over character strings. In
the following sections, we will use the the notation $\mathcal{L}(R)$ to mean
the language accepted by the regular expression $R$.

@@

\begin{lstlisting}
(* Regular expressions matching different comma and space representations *)
let comma_name = last_name . "," . wsp+ . first_name
let space_name = first_name . wsp+ . last_name

@@

\begin{prooftree}
\AxiomC{$\ell : R \Leftrightarrow S$}
\AxiomC{$R \equiv^d R'$}
\AxiomC{$S \equiv^d S'$}
\TrinaryInfC{$\ell : R' \Leftrightarrow S'$}
\end{prooftree}

@@

\begin{center}
\AxiomC{$\ell : R \Leftrightarrow S$}
\AxiomC{$R^{*!}$}
\AxiomC{$S^{*!}$}
\TrinaryInfC{$\ell^*: R \Leftrightarrow S$}
\DisplayProof
\hskip 1.5em
\AxiomC{$\ell : R \Leftrightarrow S$}
\AxiomC{$R \equiv^d R'$}
\AxiomC{$S \equiv^d S'$}
\TrinaryInfC{$\ell : R' \Leftrightarrow S'$}
\DisplayProof
\end{center}
  \caption{Bijective Lens Typing Rules}
  \label{fig:lensrules}
\end{figure}

@@

In our previous work~\cite{popl18}, we showed that given regular expressions
$R, S$ and a set of input-output pairs $\{(r_1, s_1), \ldots, (r_n, s_n)\}$, if
there is a derivation $\ell : R \Leftrightarrow S$ of a bijective lens that
maps $r_i$ to $s_i$ in the forward direction and $s_i$ to $r_i$ in the backward
direction, then there is an algorithm that computes a bijective lens $\ell' :
R' \Leftrightarrow S'$ such that $R \equiv R'$, $S \equiv S'$, $\ell'$ maps
$r_i$ to $s_i$ in the forward direction and $s_i$ to $r_i$ in the backward
direction, and $\llbracket \ell \rrbracket = \llbracket \ell'
\rrbracket$.

This implies that if there is a derivation $q : c \Leftrightarrow
c'$ of a QRE lens as well as a derivation $\ell : K(c) \Leftrightarrow K(c')$
that is consistent with the canonized examples $\{({x_1}', {y_1}'),
\ldots, ({x_n}', {y_n}')\}$, then we can synthesize a quotient lens $q$ of
type $c \Leftrightarrow c'$ that is consistent with $\{({x_1}', {y_1}'),
\ldots, ({x_n}', {y_n}')\}$. More concretely, in order to synthesize a
quotient lens $q: c \Leftrightarrow c'$, that is consistent with the examples
$\{(x_1, y_1), \ldots, (x_n, y_n)\}$ it suffices to perform the following
procedure:
\begin{enumerate}
  \item
  Compute $K(c)$, $K(c')$,
  \item
  compute $\{({x_1}', {y_1}'), \ldots, ({x_n}', {y_n}')\}$ where ${x_i}' =
  \canonizer(c)(x_i)$ and ${y_i}' = \canonizer(c')(y_i)$,
  \item
    synthesize a bijective lens $\ell : K(c) \Leftrightarrow K(c')$ that is
    consistent with the examples $\{({x_1}', {y_1}'), \ldots, ({x_n}',
    {y_n}')\}$
  \item 
  return $q = \kw{rquot}(\kw{lquot}(c, \ell), c') : c \Leftrightarrow
  c'$
\end{enumerate}
For example, recall that in the \bibtex{} to EndNote transformation we had the
QREs \cd{bibtex} and \cd{endnote} describing a \bibtex{} record and an Endnote
record respectively. We also had the example pair \cd{(bib_example ,
end_example)} where

@@

As far as synthesis goes, though there is a good deal of recent research on
synthesizing unidirectional string
transformations~\cite{singh2012learning,le-pldi-2014,gulwani-popl-2014,perelman2014test,Singh:blinkfill},
the system Optician, which we introduced in ~\cite{popl18}, is the first attempt
to synthesize bidirectional transformations that we are aware of. In that
publication, we compared Optician to two of these unidirectional string
transformers, Flash Fill~\cite{gulwani-popl-2014} and
FlashExtract~\cite{le-pldi-2014}, and found that these tools were unsuccessful
in synthesizing the complex transformatiunons that can be expressed in Optician.

@@

\bibliographystyle{plain}
\bibliography{local}

>>>>

Oftentimes, different applications store the same data in custom formats. For
example, \bibtex{} and EndNote both store bibliography data; spreadsheet data
can be represented as a CSV and as a TSV; web apis can return data as a JSON or
XML object. Synchronizing these applications amounts to synchronizing the data
formats. \emph{Lenses} comprise such synchronization functions, and the ``lens
laws'' provide restrictions on how these synchronization functions interoperate.
Bidirectional languages like Boomerang~\cite{boomerang} and
BiGul~\cite{DBLP:conf/pepm/KoZH16} use a single expression to encode all of the
lens's synchronization functions at the same time.

However, these lens laws are too restrictive for many important transformations.
In particular, these lens laws do not permit transformations that put the data
in canonical forms. In \bibtex{}, an author can be specified as \cd{lastname,
  firstname} and \cd{firstname lastname}. While it is important to allow authors
to be input in both ways, we would like these authors to only be output as
\cd{firstname lastname}, but the lens laws prevent such functions.
{\em Quotient lenses}~\cite{quotientlenses} are a generalization of lenses where
each of the data formats is quotiented by some equivalence relation and the lens
laws are taken modulo equivalence. This both permits canonization functions and
allows lenses to be simpler and more flexible. Inessential details like
whitespace and orderings of fields can be ignored when writing the
transformations between the formats.

However, while plain lenses are widely used in both standalone systems and
popular libraries, Boomerang~\cite{boomerang} is the only implementation
currently offering quotient lenses. Even though quotient lenses permit simpler
transformations between the formats, they merely push complexity elsewhere. To
write a quotient lens, programmers must write with both standard lens
combinators and \emph{canonizers}. Canonizers encode two functions, one that
maps a format to a canonical version of that format, and another that chooses
which elements of the original format correspond to values in the canonical
format. Furthermore, if a user wants to enforce that a quotient lens maps
between two different formats, they must write regular expressions specifying
such formats. In this way, writing quotient lenses involves writing 3 terms: (1)
the data formats on each side of the lens, (2) canonizers to normalize the data
and choose representatives, and (3) a lens between the data formats when in
canonical form, even though, typically, these three share a great deal of
structure. Moreover, writing the lens between the equivalence classes can be
challenging due to the rather fiddly typechecking, mainly due to
``non-ambiguity'' conditions in the lens well-formedness rules.

In this paper, we make programming quotient lenses easier in two ways. First, we
introduce a novel language of {\em Quotient Regular Expressions}, or QREs, which
enable a programmer to simultaneously specify a regular expression while
simultaneously specifying the canonical representatives on the regular
expression's language. With QREs, programmers needn't define the data and
canonization functions separately -- by adding simple annotations to the data,
the canonization functions can be inferred.

Secondly, we use QREs to define a class of quotient lenses called {\em QRE
  lenses} which map betweeen QREs, and show how to synthesize QRE lenses. Using
this tool, a lens programmer need not manually derive the lens herself, she
merely must provide a pair of QREs (describing the formats the lens should map
between), and a suite of examples (demonstrating how the lens should act on
specific instances).  The resulting lens is guaranteed to map between the
languages of the QREs and satisfy the provided examples.

In summary, our main contributions are:
\begin{enumerate}
\item We introduce a novel language of {\em Quotient Regular Expressions} (QREs)
  that provide a compact, convenient notation for regular languages and their
  canonical representatives (Section~\ref{QRE}).
\item We define a language of {\em QRE lenses}, a subset of bijective quotient
lenses whose types are given by QREs. Our main technical contribution in this
paper is a proof that each QRE lens $q$ defines the same transformation as a
QRE lens $q'$ that is in normal form. (Section~\ref{QRE-lenses}).
\item Using this normal form, we reduce the problem of {\em synthesizing}
QRE lenses to the previously studied
problem of synthesizing bijective lenses between regular languages 
(Section~\ref{synth}).
\item We extend the Boomerang {\em implementation} with QREs
and QRE lens synthesis.  We demonstrate its applicability by using it to
synthesize QRE lenses between several real-world data formats from the
{\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

\section{Quotient Regular Expressions and Lenses by Example}
\label{sec:example}
In this section, we explain the difficulties of programming with quotient lenses
and show, by way of example, how our novel language of {\em Quotient Regular
Expressions}, or QREs, and QRE lenses address these difficulties.

@@

If two \bibtex{} files differ in only ``inessential'' details, we would like
these files to map to the same EndNote representations.
First, the transformation should allow whitespace characters to occur
arbitrarily between the words in the label, author, and title fields, as well as
in between them. Second, an author's name can have the surname followed by a
comma and then the other names separated by arbitrary whitespace, or just all
the names separated by arbitrary whitespace (for example, the names ``Conway, J.
H.'' and ``J. H. Conway'' are equivalent). Lastly, the transformation should
also ignore the order in which the author and title fields occur in each
citation record.

@@

While it is possible to express the equivalence on the \bibtex{} data above in
Boomerang, actually doing so is a challenge.  Three of the inessential details
we wish to ignore---whitespace quotienting
inside fields, whitespace quotienting in between fields, and permutation
quotenting of fields---nest within each other in a way that makes them difficult
to express with the existing Boomerang combinators.

Furthermore, programming best practices require programmers provide type
annotations to their lenses to make explicit the formats the quotient lenses map
between. This practice prevents nefarious bugs that can result from the lenses
mapping between incorrect formats, but it is also verbose and time consuming.

To address these issues, we propose {\em Quotient Regular Expressions}, or QREs,
which enable the programmer to define a regular expression $R$ and its canonical
elements in the same expression. For example, with QREs we can specify \bibtex{}
records and their canonical elements, as shown in Figure~\ref{fig:example-qre}.

\begin{figure}[t]
\begin{lstlisting}
(* whitespace with " " as a canonical representative *)
let wsp = project [ \n\t\r]$^+$ $\to$ " "

(* define name representations with a space and with a comma *)
let space_name     = name | (name . wsp . (name . wsp)$^*$ . name)
let comma_name   = name | (name . "," . wsp . (name . wsp)$^*$ . name)

@@

(* squash QRE maps comma_name to space_name *)
let name = squash comma_name $\to$ space_name using (get space_to_comma)

(* define rest of bibtex fields *)
let name_sep = wsp . "and" . wsp
let bib_names  = name | (name . name_sep . (name . name_sep)$^*$ . name) 
let bib_author = "author = \"" . bib_names . "\""
let title     = word | (word . wsp . (word . wsp)$^*$ . word) 
let bib_title = "title = {" . title . "}"

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_fields  = perm (label, bib_author, bib_title) with (project ("," .wsp$^+$) $\to$ ",\n")
let bibtex    = "@book{" . bib_fields . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. }
  \label{fig:example-qre}
\end{figure}

In Figure~\ref{fig:example-qre}, the QRE \verb+project [ \n\t\r]+$^{+} \to$
\verb+" "+ matches any non-zero number of whitespace characters, with a single
space as the canonical representative. We call this QRE a \textit{projection}
because an entire regular language is projected to a single representative
string during canonization.

@@

The third example of a QRE in the Figure is \cd{bib_fields}, defined by:
\begin{center}
\begin{lstlisting}
perm (label, bib_author, bib_title) with (project ("," . wsp$^+$ ) $\to$ ",\n")
\end{lstlisting}
\end{center}

\noindent This is a \textit{permutation} QRE that specifies that the label,
title, and author fields can occur in any order, with the fields separated by a
comma followed by arbitrary amounts of whitespace. The canonical representation
is that in which the label, author and title fields occur in that order,
separated by a comma followed by a newline. The \textit{permutation} QRE is
designed for equivalences such as this, where different permutations of
equivalences are interspersed with a separator which itself has an
equivalence defined on it.

@@

Consequently, we offer the programmer some relief by identifying a class of
quotient lenses that we call {\em QRE lenses}, and show how to {\em synthesize}
QRE lenses from a pair of QREs and a set of input-output example pairs. Our key
insight is to reduce the problem of synthesizing QRE lenses to the problem of
synthesizing {\em bijective lenses}~\cite{optician}, which are bijections created
from the Boomerang lens combinators.

@@

This section introduces Quotient Regular Expressions or QREs. QREs are regular
expressions augmented with syntax that lets them simultaneously express a
language and how to convert from that language into its canonical elements.

\subsection{Syntax of Quotient Regular Expressions}
The language of Quotient Regular Expressions (QREs) is given by the following
grammar: 
\begin{align*}
c := \; &R \sep \project{R}{s} \sep \squash{c}{c'}{f} \sep
\perm{c_1, \ldots, c_n}{c} \\
  & | \; \normalize{R}{R'}{f} \sep c \semicolon c' \sep c \cdot c''' \sep (c \sep
  c') \sep c^*,
\end{align*}
where $c$ ranges over QREs, $R$ ranges over regular expressions, $f$ ranges over
functions between regular languages, and $s$ ranges over character strings. In
the following sections, we will use the the notation $\mathcal{L}(R)$ to mean
the language accepted by the regular expression $R$.

@@

\begin{lstlisting}
(* Regular expressions matching different comma and space representations *)
let comma_name = last_name . "," . wsp . first_name
let space_name = first_name . wsp . last_name

@@

\begin{prooftree}
\AxiomC{$\ell : R \Leftrightarrow S$}
\AxiomC{$R \equiv^s R'$}
\AxiomC{$S \equiv^s S'$}
\TrinaryInfC{$\ell : R' \Leftrightarrow S'$}
\end{prooftree}

@@

\begin{center}
\AxiomC{$\ell : R \Leftrightarrow S$}
\AxiomC{$R^{*!}$}
\AxiomC{$S^{*!}$}
\TrinaryInfC{$\ell^*: R \Leftrightarrow S$}
\DisplayProof
\hskip 1.5em
\AxiomC{$\ell : R \Leftrightarrow S$}
\AxiomC{$R \equiv^s R'$}
\AxiomC{$S \equiv^s S'$}
\TrinaryInfC{$\ell : R' \Leftrightarrow S'$}
\DisplayProof
\end{center}
  \caption{Bijective Lens Typing Rules}
  \label{fig:lensrules}
\end{figure}

@@

Optician~\cite{optician} showed that given regular expressions
$R, S$ and a set of input-output pairs $\{(r_1, s_1), \ldots, (r_n, s_n)\}$, if
there is a derivation $\ell : R \Leftrightarrow S$ of a bijective lens that
maps $r_i$ to $s_i$ in the forward direction and $s_i$ to $r_i$ in the backward
direction, then there is an algorithm that computes a bijective lens $\ell' :
R' \Leftrightarrow S'$ such that $R \equiv R'$, $S \equiv S'$, $\ell'$ maps
$r_i$ to $s_i$ in the forward direction and $s_i$ to $r_i$ in the backward
direction, and $\llbracket \ell \rrbracket = \llbracket \ell'
\rrbracket$.

This implies that if there is a derivation $q : c \Leftrightarrow
c'$ of a QRE lens as well as a derivation $\ell : K(c) \Leftrightarrow K(c')$
that is consistent with the canonized examples $\{({x_1}', {y_1}'),
\ldots, ({x_n}', {y_n}')\}$, then we can synthesize a quotient lens $q$ of
type $c \Leftrightarrow c'$ that is consistent with $\{({x_1}', {y_1}'),
\ldots, ({x_n}', {y_n}')\}$. More concretely, in order to synthesize a
quotient lens $q: c \Leftrightarrow c'$, that is consistent with the examples
$\{(x_1, y_1), \ldots, (x_n, y_n)\}$ it suffices to perform the following
procedure:
\begin{enumerate}
  \item
  Compute $K(c)$, $K(c')$,
  \item
  Compute $\{({x_1}', {y_1}'), \ldots, ({x_n}', {y_n}')\}$ where ${x_i}' =
  \canonizer(c)(x_i)$ and ${y_i}' = \canonizer(c')(y_i)$,
  \item
    Synthesize a bijective lens $\ell : K(c) \Leftrightarrow K(c')$ that is
    consistent with the examples $\{({x_1}', {y_1}'), \ldots, ({x_n}',
    {y_n}')\}$
  \item 
  Return $q = \kw{rquot}(\kw{lquot}(c, \ell), c') : c \Leftrightarrow
  c'$
\end{enumerate}
For example, recall that in the \bibtex{} to EndNote transformation we had the
QREs \cd{bibtex} and \cd{endnote} describing a \bibtex{} record and an Endnote
record respectively. We also had the example pair \cd{(bib_example ,
end_example)} where

@@

As far as synthesis goes, though there is a good deal of recent research on
synthesizing unidirectional string
transformations~\cite{singh2012learning,le-pldi-2014,gulwani-popl-2014,perelman2014test,Singh:blinkfill},
the system Optician, which we introduced in ~\cite{optician}, is the first attempt
to synthesize bidirectional transformations that we are aware of. In that
publication, we compared Optician to two of these unidirectional string
transformers, Flash Fill~\cite{gulwani-popl-2014} and
FlashExtract~\cite{le-pldi-2014}, and found that these tools were unsuccessful
in synthesizing the complex transformatiunons that can be expressed in Optician.

@@

\bibliographystyle{plain}
\bibliography{local,bcp}
